---
title: "BREATH API – R Vignette"
author: "GeoModelLab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)
```

## Overview

**BREATH** (Biophysical Rhythm of Ecosystem Activity & Health) simulates
hourly GPP, RECO, and NEE for deciduous broadleaf forest pixels using:

- **VPRM** (Vegetation Photosynthesis and Respiration Model) for carbon fluxes
- **SWELL** phenology for seasonal leaf dynamics
- **NASA POWER** weather data (hourly or daily disaggregated)

This vignette demonstrates the complete R workflow:

1. Submit a simulation via the REST API
2. Poll until the run completes
3. Parse the hourly CSV results
4. Compute annual carbon budgets and seasonal diagnostics
5. Visualise fluxes and SWELL phenology

### Dependencies

```{r packages}
# install.packages(c("httr2", "readr", "dplyr", "tidyr", "lubridate", "ggplot2", "patchwork"))
library(httr2)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)

BASE_URL <- "https://breath-api-thkm.onrender.com"
# For local development use: BASE_URL <- "http://localhost:5244"
```

---

## 1 · Submit a simulation

A pixel is identified by `"lat_lon"` (4 decimal places).
We simulate a deciduous broadleaf forest in the northern Apennines (Italy).

```{r submit}
pixel    <- "44.1234_10.7654"
start_yr <- 2022L
end_yr   <- 2024L

payload <- list(
  settings = list(
    pixelsRun           = list(pixel),
    startYear           = start_yr,
    endYear             = end_yr,
    inputWeather        = "hourly",        # "hourly" | "daily" | "era5land"
    modelVariant        = "Circadian",     # "Baseline" | "Pheno" | "Circadian"
    calibration         = FALSE,           # TRUE = calibrate against MODIS EVI
    calibrationVariable = "Phenology",
    simplexes           = 3L,
    iterations          = 200L,
    parametersDataFile  = "photothermalRequirements.csv"
  )
)

resp <- request(paste0(BASE_URL, "/api/breath/run")) |>
  req_method("POST") |>
  req_body_json(payload) |>
  req_timeout(30) |>
  req_perform()

resp_body_json(resp)
```

---

## 2 · Poll until complete

```{r poll}
wait_for_completion <- function(base_url, poll_interval = 10, timeout = 600) {
  deadline <- Sys.time() + timeout
  repeat {
    status <- request(paste0(base_url, "/api/breath/status")) |>
      req_timeout(10) |>
      req_perform() |>
      resp_body_json()

    cat(sprintf("\r  Status: %-12s", status$Status))

    if (status$Status == "Completed") {
      cat(sprintf("\nDone — duration: %s ms\n", status$DurationMs %||% "?"))
      return(invisible(status))
    }
    if (status$Status == "Failed")
      stop("Simulation failed. Check /api/breath/stream/logs for details.")
    if (Sys.time() > deadline)
      stop(sprintf("Simulation did not finish within %d s.", timeout))

    Sys.sleep(poll_interval)
  }
}

# Null-coalescing helper (base R < 4.4 compatible)
`%||%` <- function(a, b) if (!is.null(a)) a else b

wait_for_completion(BASE_URL)
```

---

## 3 · Download and parse results

`GET /api/results/latest` returns a CSV with one row per simulated hour.

```{r download}
csv_raw <- request(paste0(BASE_URL, "/api/results/latest")) |>
  req_timeout(60) |>
  req_perform() |>
  resp_body_string()

df <- read_csv(csv_raw, show_col_types = FALSE) |>
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date),
         doy   = yday(date))

glimpse(df)
```

---

## 4 · Annual carbon budget

Daily mean fluxes (µmol CO₂ m⁻² s⁻¹) × 86 400 s day⁻¹ × 12.01 g mol⁻¹ × 10⁻⁶ mol µmol⁻¹
= gC m⁻² day⁻¹.

```{r annual-budget}
UMOL_TO_GC_DAY <- 86400 * 12.01e-6

daily <- df |>
  group_by(date, year) |>
  summarise(across(c(GPP, RECO, NEE), mean, na.rm = TRUE), .groups = "drop") |>
  mutate(across(c(GPP, RECO, NEE), ~ .x * UMOL_TO_GC_DAY))

annual <- daily |>
  group_by(year) |>
  summarise(
    GPP_gC  = round(sum(GPP,  na.rm = TRUE), 1),
    RECO_gC = round(sum(RECO, na.rm = TRUE), 1),
    NEE_gC  = round(sum(NEE,  na.rm = TRUE), 1),
    CUE     = round(GPP_gC / RECO_gC, 3),
    Balance = if_else(NEE_gC < 0, "sink", "source"),
    .groups = "drop"
  )

knitr::kable(annual, caption = "Annual carbon budget (gC m⁻² yr⁻¹)")
```

---

## 5 · Time-series plot

```{r flux-plot, fig.height=7}
# SWELL (if present)
p_swell <- NULL
if ("SWELL" %in% names(df)) {
  swell_d <- df |>
    group_by(date) |>
    summarise(SWELL = mean(SWELL, na.rm = TRUE), .groups = "drop")

  p_swell <- ggplot(swell_d, aes(date, SWELL)) +
    geom_ribbon(aes(ymin = 0, ymax = SWELL), fill = "#facc15", alpha = 0.3) +
    geom_line(colour = "#ca8a04", linewidth = 0.7) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(y = "SWELL (EVI)", x = NULL) +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          axis.text.x = element_blank())
}

# Flux ribbons
p_flux <- daily |>
  pivot_longer(c(GPP, RECO, NEE)) |>
  ggplot(aes(date, value, colour = name)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#475569") +
  scale_colour_manual(values = c(GPP = "#4ade80", RECO = "#f87171", NEE = "#60a5fa")) +
  labs(y = "gC m⁻² day⁻¹", x = NULL, colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")

if (!is.null(p_swell)) {
  p_swell / p_flux + plot_annotation(
    title = "BREATH simulation — northern Apennines forest",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )
} else {
  p_flux
}
```

---

## 6 · Seasonal diagnostics

```{r seasonal}
seasons <- df |>
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "DJF",
    month %in% 3:5          ~ "MAM",
    month %in% 6:8          ~ "JJA",
    TRUE                    ~ "SON"
  )) |>
  group_by(year, season) |>
  summarise(
    T_mean   = round(mean(t,    na.rm = TRUE), 1),
    P_total  = round(sum(p,     na.rm = TRUE) * 24, 0),  # mm/h → mm/day, summed
    GPP_mean = round(mean(GPP,  na.rm = TRUE), 3),
    NEE_mean = round(mean(NEE,  na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  mutate(season = factor(season, levels = c("DJF","MAM","JJA","SON"))) |>
  arrange(year, season)

knitr::kable(seasons, caption = "Seasonal diagnostics")
```

---

## 7 · Phenological dates

When the `Circadian` or `Pheno` model variant is used, the output includes a
`phenophase` column that can be used to extract phenological dates (DOY).

```{r phenology}
pheno_col <- grep("phenophase", names(df), ignore.case = TRUE, value = TRUE)[1]

if (!is.na(pheno_col)) {
  pheno <- df |>
    rename(phase = all_of(pheno_col)) |>
    arrange(date) |>
    group_by(year) |>
    mutate(prev_phase = lag(phase, default = "")) |>
    filter(phase != prev_phase) |>
    summarise(
      SGS = doy[grepl("Growth",    phase, ignore.case = TRUE)][1],
      MAT = doy[grepl("Greendown", phase, ignore.case = TRUE)][1],
      SEN = doy[grepl("Senesc",    phase, ignore.case = TRUE)][1],
      EGS = doy[grepl("Dorm|Induct", phase, ignore.case = TRUE)][1],
      .groups = "drop"
    ) |>
    mutate(GSL = EGS - SGS)

  knitr::kable(pheno, caption = "Phenological dates (DOY)")
} else {
  cat("No phenophase column in output — run with modelVariant='Pheno' or 'Circadian'.\n")
}
```

---

## 8 · Multi-pixel grid run

```{r grid-run, eval=FALSE}
# 3 × 3 grid centred on the pixel, 0.1° step
lats   <- c(44.08, 44.18, 44.28)
lons   <- c(10.66, 10.76, 10.86)
pixels <- as.vector(outer(lats, lons, function(la, lo) sprintf("%.4f_%.4f", la, lo)))

grid_payload <- list(
  settings = list(
    pixelsRun           = as.list(pixels),
    startYear           = 2023L,
    endYear             = 2023L,
    inputWeather        = "hourly",
    modelVariant        = "Circadian",
    calibration         = FALSE,
    parametersDataFile  = "photothermalRequirements.csv",
    simplexes           = 3L,
    iterations          = 200L,
    calibrationVariable = "Phenology"
  )
)

request(paste0(BASE_URL, "/api/breath/run")) |>
  req_method("POST") |>
  req_body_json(grid_payload) |>
  req_perform()

wait_for_completion(BASE_URL)

csv_grid <- request(paste0(BASE_URL, "/api/results/latest")) |>
  req_perform() |>
  resp_body_string()

df_grid <- read_csv(csv_grid, show_col_types = FALSE)

# Annual NEE per pixel (gC m⁻² yr⁻¹)
nee_px <- df_grid |>
  group_by(pixel) |>
  summarise(NEE_annual = round(mean(NEE, na.rm = TRUE) * 86400 * 365 * 12.01e-6, 1)) |>
  arrange(NEE_annual)

knitr::kable(nee_px)
```

---

## 9 · Parameter overrides

Override any internal model parameter by passing a `parameterOverrides` map.
Keys follow the `className_propertyName` convention.

```{r overrides, eval=FALSE}
tuned_payload <- list(
  settings = list(
    pixelsRun           = list(pixel),
    startYear           = 2023L,
    endYear             = 2023L,
    inputWeather        = "hourly",
    modelVariant        = "Circadian",
    calibration         = FALSE,
    parametersDataFile  = "photothermalRequirements.csv",
    simplexes           = 3L,
    iterations          = 200L,
    calibrationVariable = "Phenology"
  ),
  parameterOverrides = list(
    parPhotosynthesis_maximumQuantumYieldOver = 0.08,  # α overstory (default 0.06)
    parPhotosynthesis_halfSaturationTree      = 350,   # k½ PAR (default 400)
    parRespiration_referenceRespiration       = 1.2    # R base (default 1.5)
  )
)

request(paste0(BASE_URL, "/api/breath/run")) |>
  req_method("POST") |>
  req_body_json(tuned_payload) |>
  req_perform()

wait_for_completion(BASE_URL)
```

---

## 10 · Custom weather data

By default BREATH downloads meteorological inputs automatically from NASA POWER.
If you have your own station data, reanalysis output, or ERA5 downloads you can
upload a daily weather CSV and the next simulation for that pixel will use it
instead, skipping the remote download entirely.

### 10.1 · Required CSV format

The file must be comma-separated with the following columns (invariant culture,
dot as decimal separator):

| Column | Unit | Notes |
|--------|------|-------|
| `Date` | yyyy-MM-dd | One row per day |
| `PAR` | MJ m⁻² d⁻¹ | Photosynthetically active radiation |
| `airTemperatureMaximum` | °C | Daily maximum |
| `airTemperatureMinimum` | °C | Daily minimum |
| `solarRadiation` | MJ m⁻² d⁻¹ | Global shortwave |
| `dewPointTemperature` | °C | Used to compute VPD |
| `precipitation` | mm d⁻¹ | Total daily rainfall |
| `latitude` | ° | Pixel latitude (constant per file) |

Retrieve a minimal template with:

```{r weather_template, eval=FALSE}
template_csv <- request(paste0(BASE_URL, "/api/weather/template")) |>
  req_perform() |>
  resp_body_string()

cat(template_csv)
```

### 10.2 · Upload

```{r weather_upload, eval=FALSE}
# Build a minimal daily data frame for demonstration
library(readr)

pixel_custom <- "44.6813_11.0217"

weather_df <- tibble::tibble(
  Date                    = seq(as.Date("2022-01-01"), as.Date("2024-12-31"), by = "day"),
  PAR                     = 5.0,           # replace with real values
  airTemperatureMaximum   = 12.0,
  airTemperatureMinimum   =  2.0,
  solarRadiation          = 10.0,
  dewPointTemperature     = -1.0,
  precipitation           =  1.5,
  latitude                = 44.6813
)

weather_csv <- format_csv(weather_df)

# Upload — the file is stored server-side as {pixel}_weather.csv
resp_upload <- request(paste0(BASE_URL, "/api/weather/upload")) |>
  req_url_query(pixel = pixel_custom) |>
  req_method("POST") |>
  req_headers("Content-Type" = "text/csv") |>
  req_body_raw(weather_csv, type = "text/csv") |>
  req_perform()

resp_upload |> resp_body_json()
```

### 10.3 · Run with custom weather

After uploading, run the simulation as usual.
BREATH detects the cached file and skips the NASA POWER call automatically.

```{r weather_run, eval=FALSE}
custom_payload <- list(
  settings = list(
    pixelsRun           = list(pixel_custom),
    startYear           = 2022L,
    endYear             = 2024L,
    inputWeather        = "daily",   # must match the file's daily resolution
    modelVariant        = "Pheno",
    calibration         = FALSE,
    parametersDataFile  = "photothermalRequirements.csv",
    simplexes           = 3L,
    iterations          = 200L,
    calibrationVariable = "Phenology"
  )
)

request(paste0(BASE_URL, "/api/breath/run")) |>
  req_method("POST") |>
  req_body_json(custom_payload) |>
  req_perform()

wait_for_completion(BASE_URL)
```

### 10.4 · List and delete cached weather files

```{r weather_manage, eval=FALSE}
# List all pixels with a cached weather file
request(paste0(BASE_URL, "/api/weather/list")) |>
  req_perform() |>
  resp_body_json()

# Delete a specific pixel's file to revert to NASA POWER
request(paste0(BASE_URL, "/api/weather/", pixel_custom)) |>
  req_method("DELETE") |>
  req_perform() |>
  resp_body_json()
```

---

## Session info

```{r session}
sessionInfo()
```
